---
title: "ordination plots"
author: "jen nguyen"
date: "10/8/2021"
output: html_document
---

Overarching question in this ASV level analysis:
- Are quantifications of distance between paired mucus and tissue communities greater in harder than gentler scraping?

Methods
    1. Bray-Curtis distance by NMDS
    2. Bray-Curtis distance by PCoA

Plots
    a. ordination of luminal, mucus and tissue samples
    b. ordination of mucus and tissue samples only
    
Results:    
    1. luminal samples cluster together, but mucus and tissue are not clearly separable.
    2. no clear separation between gentle and hard scrapings. such is less expected given similarity between mucus and tissue.


```{r initialize, include=FALSE}
library(ggplot2)
library(here)
library(phyloseq)
library(Biostrings)
library(tidyverse)
theme_set(theme_bw())
```

```{r read sample metadata into data frame}
meta <- read_csv(here("data","meta","sample_metadata.csv"),
                      col_types = cols(
                        Name = col_character(),
                        Extraction = col_date(format = ""),
                        Mouse = col_character(),
                        Replicate = col_character(),
                        Location = col_character()#,
                        #Pair = col_double()
                      ))
df <- meta %>% column_to_rownames("Name")
```

```{r load dada2 outputs}
seqtab <- readRDS(here("output","dada2","seqtable.rds")) # raw sequence table
seqtab.nochim <- readRDS(here("output","dada2","seqtable_nochimeras.rds")) # sequence table w/o chimeras
taxa <- readRDS(here("output","dada2","taxonomic_assignments.rds")) # taxonomic assignments
```

```{r construct phyloseq object from dada2 output}
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(df), 
               tax_table(taxa))
```

```{r prune out mock, water and extraction controls}
ps <- prune_samples(sample_names(ps) != "JN18", ps) # Remove extraction control
ps <- prune_samples(sample_names(ps) != "mock-", ps) # Remove mock sample
ps <- prune_samples(sample_names(ps) != "mock-2", ps) # Remove mock sample
ps <- prune_samples(sample_names(ps) != "NTC", ps) # Remove water control
```

```{r ASV short names}
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps
```


```{r 1.1 Bray Curtis, part 1: lumen, mucus, tissue}
# Transform data to proportions as appropriate for Bray-Curtis distances
ps.prop <- transform_sample_counts(ps, function(otu) otu/sum(otu))
ord.nmds.bray <- ordinate(ps.prop, method="NMDS", distance="bray")

# Bray-Curtis ordination
plot_ordination(ps.prop, ord.nmds.bray, color="Location", shape = "Mouse", title="Bray NMDS, 1st seq")+
  geom_point(size=2)
```

```{r 1.2 Bray Curtis, part 2: mucus and tissue only}
ps2 <- prune_samples(sample_names(ps) != "JN01", ps) # Remove luminal sample
ps2 <- prune_samples(sample_names(ps2) != "JN04", ps2) # Remove luminal sample
ps2 <- prune_samples(sample_names(ps2) != "JN07", ps2) # Remove luminal sample

# Transform data to proportions as appropriate for Bray-Curtis distances
ps.prop2 <- transform_sample_counts(ps2, function(otu) otu/sum(otu))
ord.nmds.bray2 <- ordinate(ps.prop2, method="NMDS", distance="bray")

# Bray-Curtis ordination
plot_ordination(ps.prop2, ord.nmds.bray2, color="Location", shape = "Mouse", title="Bray NMDS, 1st seq")+
  geom_point(size=2)
```


```{r 1.3 Bray Curtis, part 3: mucus and tissue only, identifying gentle vs hard}
plot_ordination(ps.prop2, ord.nmds.bray2, color ="Extraction", shape = "Replicate",  title="Bray NMDS, 1st seq") +
  geom_point(size=2)
```


```{r 2.1 Bray Curtis, PCoA: lumen, mucus, tissue}
ord.nmds.bray3 <- ordinate(ps.prop, method="PCoA", distance="bray")

# Bray-Curtis ordination
plot_ordination(ps.prop, ord.nmds.bray3, color="Location", shape = "Mouse", title="Bray PCoA, 1st seq")+
  geom_point(size=2)
```

```{r 2.2 Bray Curtis, PCoA: mucus and tissue only}
ord.nmds.bray4 <- ordinate(ps.prop2, method="PCoA", distance="bray")

# Bray-Curtis ordination
plot_ordination(ps.prop2, ord.nmds.bray4, color="Location", shape = "Mouse", title="Bray PCoA, 1st seq")+
  geom_point(size=2)
```

```{r 2.3 Bray Curtis, PCoA: mucus and tissue only, identifying gentle vs hard}
plot_ordination(ps.prop2, ord.nmds.bray4, color ="Extraction", shape = "Replicate",  title="Bray PCoA, 1st seq") +
  geom_point(size=2)
```