---
title: "dada2 analysis for blue-jam project: revised for library details"
author: "jen nguyen"
date: "10/29/2021"
output: html_notebook
---

Samples collected on 2021-07-23 and 2021-07-30 were re-sequenced by Gut4Health (different library prep from first sequencing run). This R Markdown file is the revised analysis of the 2nd sequencing run, received by Jen on October 7, 2021.

This 2nd run used the this Zymo library prep kit:
https://www.zymoresearch.com/products/quick-16s-plus-ngs-library-prep-kit

This prep kit uses these primers, which are sequenced and included in the reads:
16S V3-V4 Primer Sequences (adapters not included) – 341f (CCTACGGGDGGCWGCAG, CCTAYGGGGYGCWGCAG, 17 bp) and 806r (GACTACNVGGGTMTCTAATCC, 24 bp). The forward primer 341f is a mixture of the two sequences listed.

Expected lengths of sequenced amplicons (primers + 16S region) from 16S 1 of each bacterial species in mock:
  - B. subtilis = 466 bp
  - E. faecalis = 465 bp
  - E. coli = 465 bp
  - L. fermentum = 465 bp
  - L. monocytogenes = 465 bp
  - P. aeruginosa = 465 bp
  - S. enteria = 465 bp
  - S. aureus = 465 bp


This dada2 analysis:
  1. Trims the beginning 17 bp of fwd reads and 24 bp of rev reads (primer lengths).
  2. Truncates fwd reads at 277 bp.
  3. Truncates rev reads at 250 bp.
  4. Requires 12 bp overlap to merge reads.

Total length 466 - fwd primer 17 - rev primer 24 = 425 bp
Truncated fwd read 277 + rev read 250 = 527 bp - fwd primer 17 - rev primer 24 = 486


```{r import libraries, include=FALSE}
library(dada2); packageVersion("dada2")
library(ggplot2)
library(here)
library(phyloseq)
library(Biostrings)
library(tidyverse)
library(fs)
```

```{r define paths to seq data}
path_raw <- "/Users/jen/Documents/TropiniLab/Data/Sequencing/2021-10-07/raw"
path_filtered <- "/Users/jen/Documents/TropiniLab/Data/Sequencing/2021-10-07/filtered_oct29"
list.files(path_raw)
```

```{r format forward and reverse fastq filenames}
# SAMPLENAME_R1_001.fastq and SAMPLENAME_R2_001.fastq
fnFs <- sort(list.files(path_raw, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path_raw, pattern="_R2_001.fastq", full.names = TRUE))
fnFs %>% path_file %>% head
```

```{r extract sample names}
# assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
print(sample.names)
```

```{r check fwd reads of first two samples}
plotQualityProfile(fnFs[1:2]) +
  geom_vline(xintercept = c(17,277)) # trim & truncation length of fwd reads
```
```{r check rev reads of first two samples}
plotQualityProfile(fnRs[1:2]) +
  geom_vline(xintercept = c(24,250)) # trim & truncation length of rev reads
```

```{r prepare to place filtered files in filtered/ subdirectory}
filtFs <- file.path(path_filtered, paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path_filtered, paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

```{r place filtered files in filtered/ subdirectory}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(277,250),
              maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=TRUE, trimLeft=c(17,24)) # On Windows set multithread=FALSE
head(out)
```

```{r learn the error rates and perform sample inference}
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
plotErrors(errF, nominalQ=TRUE)

dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
dadaFs[[1]]
```

```{r merge paired reads}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
```

```{r construct sequence table}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```


```{r inspect distribution of sequence lengths}
table(nchar(getSequences(seqtab)))
```


```{r remove chimeras}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
```

```{r calculate fraction of reads that were chimeras}
sum(seqtab.nochim)/sum(seqtab)
```

```{r track reads through the pipeline}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)
write.csv(track, here::here('output', 'dada2', 'track_oct29.csv'))
```

```{r locate taxonomy database files}
"/Users/jen/databases" %>% list.files()
```


```{r assign taxonomy using the training set}
taxa <- assignTaxonomy(
  seqtab.nochim, 
  "/Users/jen/databases/silva_nr99_v138.1_train_set.fa.gz",
  multithread=TRUE
)
```

```{r add species}
taxa <- addSpecies(
  taxa, 
  "/Users/jen/databases/silva_species_assignment_v138.1.fa.gz",
  allowMultiple = TRUE
)
```

```{r inspect taxonomic assignments}
taxa.print <- taxa # Removing sequence row names for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

```{r evaluate accuracy, mock}
unqs.mock <- seqtab.nochim["mock2",]
unqs.mock <- sort(unqs.mock[unqs.mock>0], decreasing=TRUE) # Drop ASVs absent in the Mock
cat("DADA2 inferred", length(unqs.mock), "sample sequences present in the Mock community.\n")
```
The ZymoBIOMICS Gut Microbiome Standard has 21 species!
Less truncation gave 21 ASVs in Mock Community whereas more truncation gives 29!

```{r evaluate accuracy, water control}
unqs.water <- seqtab.nochim["NTC2",]
unqs.water <- sort(unqs.mock[unqs.water>0], decreasing=TRUE) # Drop ASVs absent in the Mock
cat("DADA2 inferred", length(unqs.water), "sample sequences present in the water control\n")
```

Hm, how to interpret this?

More ASVs than species in the mock community can arise by having multiple 16S sequences per species. This is something to check from the sequences available from the mock communities.

More ASVs can also arise via contamination. If there is a sequence in the mock as well as in the samples, then this is good evidence of contamination.

```{r, eval = FALSE}
mock.ref <- getSequences(file.path(path, "HMP_MOCK.v35.fasta"))
match.ref <- sum(sapply(names(unqs.mock), function(x) any(grepl(x, mock.ref))))
cat("Of those,", sum(match.ref), "were exact matches to the expected reference sequences.\n")
```

Skipped checking that mock community sequences were exact matches to expected reference sequences because the files associated with our community are currently available in a different format.

```{r saving DADA2 output}
saveRDS(seqtab.nochim,file=here("output","dada2_20211029","seqtable_nochimeras_oct29.rds"))
saveRDS(seqtab,file=here("output","dada2_20211029","seqtable_oct29.rds"))
saveRDS(taxa,file=here("output","dada2_20211029","taxonomic_assignments_oct29.rds"))
saveRDS(errF,file=here("output","dada2_20211029","errF_oct29.rds"))
saveRDS(errR,file=here("output","dada2_20211029","errR_oct29.rds"))
saveRDS(dadaFs,file=here("output","dada2_20211029","dadaFs_oct29.rds"))
saveRDS(dadaRs,file=here("output","dada2_20211029","dadaRs_oct29.rds"))
```


```{r}
sessionInfo()
```

