---
title: "troubleshooting merge step of dada2 analysis for 2nd seq run data"
author: "jen nguyen"
date: "10/26/2021"
output: html_notebook
---

Testing analysis improvements on a subset of samples from the following:
- collection: 2021-07-23 and 2021-07-30 
- sequencing run: second by Gut4Health on the same submitted DNA samples
- data: different library prep and sequencer than first run, fastq.gz files received by Jen on October 7, 2021


This code is the 1st of 3 tests that will systematically change the initial analysis Jen performed on this data in 2021-10-14-dada2.Rmd. The 3 changes are as follows:
  
  minidada-1: trimming first 20 bp of raw reads
  minidada-2: test1 + truncating less to accomodate the lost 40 bp from trimming
  minidada-3: test1 + test2 + relaxing required overlap to 12 bp to merge


Chosen sample subset:
  1. mock2 (known composition and clearly erroneous in initial analysis)
  2. Jen18 (negative extraction control)
  3. NTC2  (negative water control)
  4. Jen01 (luminal sample, drops from 146,185 to 39,688 reads after merge step)
  5. Jen02 (mucus sample, drops from 218,523 to 190,111 reads after merge step)
  6. Jen03 (tissue sample, drops from 230,981 to 138,234 reads after merge step)
  

```{r import libraries, include=FALSE}
library(dada2); packageVersion("dada2")
library(ggplot2)
library(here)
library(phyloseq)
library(Biostrings)
library(tidyverse)
library(fs)
```

```{r define paths to data}
path_raw <- "/Users/jen/Documents/TropiniLab/Data/Sequencing/2021-10-07/testset/raw"
path_filtered <- "/Users/jen/Documents/TropiniLab/Data/Sequencing/2021-10-07/testset/filtered_1"
list.files(path_raw)
```

```{r id fwd and rev files}
# Forward and reverse fastq filenames have format: SAMPLENAME_R1_001.fastq and SAMPLENAME_R2_001.fastq
fnFs <- sort(list.files(path_raw, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path_raw, pattern="_R2_001.fastq", full.names = TRUE))
fnFs %>% path_file %>% head
```

```{r extract sample names}
# assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
print(sample.names)
```

```{r check fwd reads of first two samples}
# initial truncation length of fwd reads was at 248 bp
plotQualityProfile(fnFs[1:2]) +
  geom_vline(xintercept = 248)
```
```{r check rev reads of first two samples}
# initial truncation length of rev reads was at 225 bp
plotQualityProfile(fnRs[1:2]) +
  geom_vline(xintercept = 225)
```

```{r prepare to place filtered files in filtered/ subdirectory}
filtFs <- file.path(path_filtered, paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path_filtered, paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

```{r place filtered files in filtered/ subdirectory}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(248,225),
              maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=TRUE, trimLeft=c(20,20)) # On Windows set multithread=FALSE
head(out)
```

```{r learn the error rates, fwd}
errF <- learnErrors(filtFs, multithread=TRUE)
```
```{r learn the error rates, rev}
errR <- learnErrors(filtRs, multithread=TRUE)
```


```{r plot learned error rates}
plotErrors(errF, nominalQ=TRUE)
```

```{r sample inference, fwd}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
```

```{r sample inference, rev}
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```

```{r}
dadaFs[[1]]
```


```{r merge paired reads}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
```

```{r construct sequence table}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```


```{r inspect distribution of sequence lengths}
table(nchar(getSequences(seqtab)))
```


```{r remove chimeras}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
```

```{r calculate fraction of reads that were chimeras}
sum(seqtab.nochim)/sum(seqtab)
```
Chimeras make up a substantial portion (~15%) of merged sequence variants, but are only about 1% of the merge sequenced reads when accounting for the abundance of these variants.

```{r track reads through the pipeline}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)
write.csv(track, here::here('output', 'minidada_1', 'track_1.csv'))
```


```{r locate taxonomy database files}
"/Users/jen/databases" %>% list.files()
```


```{r assign taxonomy using the training set}
taxa <- assignTaxonomy(
  seqtab.nochim, 
  "/Users/jen/databases/silva_nr99_v138.1_train_set.fa.gz",
  multithread=TRUE
)
```

```{r add species}
taxa <- addSpecies(
  taxa, 
  "/Users/jen/databases/silva_species_assignment_v138.1.fa.gz",
  allowMultiple = TRUE
)
```

```{r inspect taxonomic assignments}
taxa.print <- taxa # Removing sequence row names for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

```{r evaluate accuracy, mock}
unqs.mock <- seqtab.nochim["mock2",]
unqs.mock <- sort(unqs.mock[unqs.mock>0], decreasing=TRUE) # Drop ASVs absent in the Mock
cat("DADA2 inferred", length(unqs.mock), "sample sequences present in the Mock community.\n")
```
The ZymoBIOMICS Gut Microbiome Standard has 21 species!

```{r evaluate accuracy, water control}
unqs.water <- seqtab.nochim["NTC2",]
unqs.water <- sort(unqs.mock[unqs.water>0], decreasing=TRUE) # Drop ASVs absent in the Mock
cat("DADA2 inferred", length(unqs.water), "sample sequences present in the water control\n")
```


```{r, eval = FALSE}
mock.ref <- getSequences(file.path(path, "HMP_MOCK.v35.fasta"))
match.ref <- sum(sapply(names(unqs.mock), function(x) any(grepl(x, mock.ref))))
cat("Of those,", sum(match.ref), "were exact matches to the expected reference sequences.\n")
```

Skipped checking that mock community sequences were exact matches to expected reference sequences because the files associated with our community are currently available in a different format.

```{r saving DADA2 output}
saveRDS(seqtab.nochim,file=here("output","minidada_1","seqtable_nochimeras_1.rds"))
saveRDS(seqtab,file=here("output","minidada_1","seqtable_1.rds"))
saveRDS(taxa,file=here("output","minidada_1","taxonomic_assignments_1.rds"))
saveRDS(errF,file=here("output","minidada_1","errF_1.rds"))
saveRDS(errR,file=here("output","minidada_1","errR_1.rds"))
saveRDS(dadaFs,file=here("output","minidada_1","dadaFs_1.rds"))
saveRDS(dadaRs,file=here("output","minidada_1","dadaRs_1.rds"))
```


```{r}
sessionInfo()
```

