---
title: "ordination plots for re-sequenced data received Nov 22, re-analysed with dada2 on Dec 06"
author: "jen nguyen"
date: "12/06/2021"
output: html_document
---

Overarching question in this ASV level analysis:
- Are quantifications of distance between paired mucus and tissue communities greater in harder than gentler scraping?

Methods
    1. Bray-Curtis distance by NMDS
    2. Bray-Curtis distance by PCoA

Plots
    a. ordination of luminal, mucus and tissue samples
    b. ordination of mucus and tissue samples only
    
Results:    
    1. luminal samples cluster together, but mucus and tissue are not clearly separable.
    2. no clear separation between gentle and hard scrapings. such is less expected given similarity between mucus and tissue.


```{r initialize, include=FALSE}
library(ggplot2)
library(here)
library(phyloseq)
library(Biostrings)
library(tidyverse)
theme_set(theme_bw())
```

```{r read sample metadata into data frame}
meta <- read_csv(here("data","meta","sample_metadata_nov22.csv"),
                      col_types = cols(
                        Name = col_character(),
                        Extraction = col_date(format = ""),
                        Mouse = col_character(),
                        Replicate = col_character(),
                        Location = col_character()#,
                        #Pair = col_double()
                      ))
df <- meta %>% column_to_rownames("Name")
```

```{r load dada2 outputs specific to Nov 22 data & Dec 06 analysis}
seqtab <- readRDS(here("output","dada2_run3","seqtable_dec06.rds")) # raw sequence table
seqtab.nochim <- readRDS(here("output","dada2_run3","seqtable_nochimeras_dec06.rds")) # seq table w/o chimeras
taxa <- readRDS(here("output","dada2_run3","taxonomic_assignments_dec06.rds")) # taxonomic assignments
```

```{r construct phyloseq object from dada2 output}
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(df), 
               tax_table(taxa))
```

```{r prune out mock, water and extraction controls}
ps <- prune_samples(sample_names(ps) != "JenUn18", ps) # Remove extraction control
ps <- prune_samples(sample_names(ps) != "mock1", ps) # Remove mock sample
ps <- prune_samples(sample_names(ps) != "mock2", ps) # Remove mock sample
ps <- prune_samples(sample_names(ps) != "NTC", ps) # Remove water control
```

```{r ASV short names}
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps
```


```{r 1.1 Bray Curtis, part 1: lumen, mucus, tissue}
# Transform data to proportions as appropriate for Bray-Curtis distances
ps.prop <- transform_sample_counts(ps, function(otu) otu/sum(otu))
ord.nmds.bray <- ordinate(ps.prop, method="NMDS", distance="bray")

# Bray-Curtis ordination
plot_ordination(ps.prop, ord.nmds.bray, color="Location", shape = "Mouse", title="Bray NMDS, 3rd seq (Dec 06)")+
  geom_point(size=2)
```

```{r 1.2 Bray Curtis, part 2: mucus and tissue only}
ps2 <- prune_samples(sample_names(ps) != "JenUn01", ps) # Remove luminal sample
ps2 <- prune_samples(sample_names(ps2) != "JenUn04", ps2) # Remove luminal sample
ps2 <- prune_samples(sample_names(ps2) != "JenUn07", ps2) # Remove luminal sample

# Transform data to proportions as appropriate for Bray-Curtis distances
ps.prop2 <- transform_sample_counts(ps2, function(otu) otu/sum(otu))
ord.nmds.bray2 <- ordinate(ps.prop2, method="NMDS", distance="bray")

# Bray-Curtis ordination
plot_ordination(ps.prop2, ord.nmds.bray2, color="Location", shape = "Mouse", title="Bray NMDS, 3rd seq (Dec 06)")+
  geom_point(size=2)
```


```{r 1.3 Bray Curtis, part 3: mucus and tissue only, identifying gentle vs hard}
plot_ordination(ps.prop, ord.nmds.bray, color ="Extraction", shape = "Replicate",  title="Bray NMDS, 3rd seq (Dec 06)") +
  geom_point(size=2)
```


```{r 2.1 Bray Curtis, PCoA: lumen, mucus, tissue}
ord.nmds.bray3 <- ordinate(ps.prop, method="PCoA", distance="bray")

# Bray-Curtis ordination
plot_ordination(ps.prop, ord.nmds.bray3, color="Location", shape = "Mouse", title="Bray PCoA, 3rd seq (Dec 06)")+
  geom_point(size=2)
```

```{r 2.2 Bray Curtis, PCoA: mucus and tissue only}
ord.nmds.bray4 <- ordinate(ps.prop2, method="PCoA", distance="bray")

# Bray-Curtis ordination
plot_ordination(ps.prop2, ord.nmds.bray4, color="Location", shape = "Mouse", title="Bray PCoA, 3rd seq (Dec 06)")+
  geom_point(size=2)
```

```{r 2.3 Bray Curtis, PCoA: mucus and tissue only, identifying gentle vs hard}
plot_ordination(ps.prop2, ord.nmds.bray4, color ="Extraction", shape = "Replicate",  title="Bray PCoA, 3rd seq (Dec 06)") +
  geom_point(size=2)
```