---
title: "ddpcr"
author: "jen nguyen"
date: "9/9/2021"
output: html_document
---

```{r initialize}
library(ggplot2)
library(here)
library(phyloseq)
library(Biostrings)
library(tidyverse)
#theme_set(theme_bw())
```

```{r read sample metadata into data frame}
meta <- read_csv(here("data","meta","sample_metadata.csv"),
                      col_types = cols(
                        Name = col_character(),
                        Extraction = col_date(format = ""),
                        Mouse = col_double(),
                        Replicate = col_double(),
                        Location = col_character(),
                      ))
df <- meta %>% column_to_rownames("Name") # data frame with same structure as ben's example
```

```{r read ddPCR data}
d0 <- read_csv(here("data","ddPCR","G4H_data_ddPCR_20210812.csv"),
                   col_types = cols(
                     Well = col_character(),
                     Sample = col_character(),
                     Dilution = col_double(),
                     Conc = col_double(),
                     Original_conc = col_double(),
                     Status = col_character(),
                     Supermix = col_character(),
                     Accepted_droplets = col_double(),
                     Positives = col_double(),
                     Negatives = col_double(),
                     Location = col_character(),
                     Mouse = col_double(),
                     Replicate = col_double(),
                     Pair = col_double(),
                     Method = col_character()
                   ))
```

General Questions:

1. how does 16S copy # vary across sample location (lumen, mucus, tissue)?
2. how variable is total 16S copy # of mucus + tissue between gut segments? how variable is it between mice?
3. is total 16S copy # of mucus + tissue correlated with 16S copy # from lumen?


Questions that get at whether gentle (mucus & tissue samples <10) or harder scraping makes a difference in how we measure mucus and tissue communities:

4. is microbial load (as estimated by ddPCR) different between mucus and tissue?
5. what proportion of total 16S copy # of mucus + tissue is from mucus?

```{r question 1}
# 1. how does 16S copy number vary across sample location (lumen, mucus, tissue)?
d0 %>%
  filter(Location != 'Water') %>%
  ggplot(aes(Sample, Conc, color = Location)) +
  geom_point() +
  facet_wrap(~Mouse, scales = 'free_x')
```

```{r question 2a}
# 2. how variable is total 16S copy number of mucus + tissue between gut segments? how variable is it between mice?
d0 %>% 
  ggplot(aes(x = Pair, y = Conc, fill = Location)) +
  geom_col()
```
Pair 1 = mouse 1, first cm (gentle)
Pair 2 = mouse 2, first cm (gentle)
Pair 3 = mouse 3, first cm (gentle)
Pair 4 = mouse 3, second cm (gentle)
Pair 5 = mouse 4, first cm (harder)
Pair 6 = mouse 5, first cm (harder)
Pair 7 = mouse 6, first cm (harder)

The total number of reads (mucus + tissue) is not substantially different between gentle and harder scrapings, though seems to trend higher in harder scraped samples.

Proportion of mucus reads appears increased in harder scrapings, and the proportion of reads from tissue decreased. Could we be taking tissue associated bacteria from T?
It seems that mucus should be more bacteria dense, and therefore it may rather be that we are more fully collecting mucus with harder scrapings.

This latter hypothesis would be supported by the observation that the taxonomic composition of mucus vs. tissue are more distinct in harder scrapings.


```{r question 2b}
# 2. how variable is total 16S copy number of mucus + tissue between gut segments and/or type of scraping?
d1 <- d0 %>%
  with_groups(Pair, mutate, S = sum(Conc))          # S = dna conc from mucus (M) + tissue (T) for each gut segment

d1 %>% 
  filter(Location %in% c('Mucus', 'Tissue')) %>%
  ggplot(aes(x = Method, y = S, color = Method)) +
  geom_point()
```
```{r question 2c}
# 2. how variable is total 16S copy number of mucus + tissue between mice?
d1 %>% 
  filter(Location %in% c('Mucus', 'Tissue')) %>%
  ggplot(aes(x = Mouse, y = S, color = Method)) +
  geom_point()
```


```{r question 3}
# 3. is total 16S copy conc of mucus + tissue correlated with 16S copy concentration from lumen?
d2g <- d1 %>% 
  filter(Replicate != 1) %>%  # NOOO. this removes pair 4 altogether
  with_groups(Mouse, mutate, Luminal_conc = sum(Conc)-S)

d2g[,17][d2g[,17] == 0] <- NA

d2g %>% 
  filter(Location %in% c('Mucus', 'Tissue')) %>%
  ggplot(aes(x = Luminal_conc, y = Conc, color = Location)) +
  geom_point() +
  geom_line(aes(group = Location))

```
Positive correlation between luminal concentration from gut segment (or nearby) and 
the 16S concentration measured from mucus and tissue with gentle scraping.

The only mucus + tissue pair associated with a luminal measurement and collected with a hard scraping
was JN10 and JN11. (Associated luminal sample was JN07.)
Luminal = 2333.74
Mucus = 464.93
Tissue = 183.64


```{r question 3, eval=FALSE, include=FALSE}
# 3. is total 16S copy conc of mucus + tissue correlated with 16S copy concentration from lumen?

hloc <- c('Luminal', 'Mucus', 'Tissue')
hconc <- c(2333.74, 464.93, 183.64)
hlumc <- c(2333.74, 2333.74, 2333.74)

d2h <- data.frame(hloc, hconc, hlumc)
d2h %>% 
  filter(hloc %in% c('Mucus', 'Tissue')) %>% 
  ggplot(aes(x = hlumc, y = hconc, color = hloc)) +
  geom_point()

```
^still learning how to overlay data from different dataframes using ggplot....


```{r question 4}
# 4. is microbial load (as estimated by ddPCR) different between mucus and tissue?
# 5. what proportion of total 16S copy # of mucus + tissue is from mucus?

d1 <- dddata %>%
  with_groups(Pair, mutate, Fraction_of_S = Conc / sum(Conc)) 

d1 %>%
  ggplot(aes(x = Pair, y = Fraction_of_S, fill = Location)) +
  geom_col()
```
Harder scraping technique increases proportion of total reads in mucus. This increase in M can result from either:
1. harder scraping collects more mucus and more bacteria
2. harder scraping collects some tissue-associated bugs


How to distinguish between these two scenarios?

Mucus should be more bacteria dense (REF), and therefore it may rather be that we are more fully collecting mucus with harder scrapings.

This latter hypothesis would be supported by the observation that the taxonomic composition of mucus vs. tissue are more distinct in harder scrapings.

Crypt-specific core microbiota of mouse colon (Pédron et al. 2012)
Abundant in crypts:
1. Proteobacteria (Phyl); Moraxellaceae (Fam); Acinetobacter (Genus): 40-50% of crypt microbiome vs. 2% in lumen
2. Proteobacteria (Phyl); Comamonadaceae (Fam); Comomonas (Genus): 2-8% of crypt microbiome
3. Proteobacteria (Phyl); Xanthomonadaceae (Fam); Stenotrophomonas (Genus): 1-30% of crypt microbiome

Not in crypts, abundant in lumen:
1. Firmicutes = majority of luminal sequences, basically absent from crypts
