---
title: "phyloseq analysis for blue-jam project (ASV level)"
author: "jen nguyen"
date: "9/7/2021"
output: html_document
---
Starting to learn how to work with this data in R!

Current state of analysis:

1. Most of our samples fell far below the read count threshold (10,000 reads) considered to be good sequencing depth by Gut4Heath.
   - Gut4Health will re-run our samples using a different library prep protocol in effort to gain more reads per sample. 
   - ddPCR counts less 16S copies per volume of mucus sample in gentle scrapings than in harder scrapings. Counts are more compatible in harder scrapings.
   - How to interpret this? Assume total mucus + tissue is fixed per gut segment and model/draw out what will happen under different scenarios.
   
2. With current data, some differences in proportional abundance between mucus vs. tissue communities can be observed.
   - I can identify which bugs change in proportional abundance between mucus and tissue.
   - Next steps: cross-check these identifications with literature on mucosal and crypt microbiome, are these consistent? 
   - Is one prep (gentle or hard scraping) more reasonable than another?

3. Alpha diversity metrics might suggest that the hard scraping is the way to go.
   - Reduced Shannon diversity in paired mucus and tissue samples gathered from hard scraping could mean we are separating mucus vs tissue specific bugs from one another. This hypothesis would be supported if pooling paired mucus and tissue samples serves to increase diversity.
   - By Simpson diversity, tissue diversity seems more constant while mucus diversity drops in hard scraping samples. I don't really understand this yet.
   - Next step: count # ASVs and evenness to describe.

4. Bray-curtis doesn't yet suggest one way to go. Samples from mucus vs. tissue don't seem to cluster more closely to same location. Rather it might be that the intestinal segment and/or mouse more strongly determines distance between sample communities.
   - Next step: edit this analysis to determine whether this latter hypothesis could be true.
   
5. Everything so far is Family level. Consider using ASV level classification with thresholds to determine what is interesting, what to look at. Then continue going up levels to determine what level of resolution is needed to describe the process. Example: Carolina was finding that a lot of bugs went extinct, but eventually went higher and higher to ultimately say S24-7 family went extinct.

```{r initialize}
library(ggplot2)
library(here)
library(phyloseq)
library(Biostrings)
library(tidyverse)
theme_set(theme_bw())
```

```{r read sample metadata into data frame}
meta <- read_csv(here("data","meta","sample_metadata.csv"),
                      col_types = cols(
                        Name = col_character(),
                        Extraction = col_date(format = ""),
                        Mouse = col_double(),
                        Replicate = col_double(),
                        Location = col_character(),
                      ))
df <- meta %>% column_to_rownames("Name") # data frame with same structure as ben's example
```

```{r load dada2 outputs}
seqtab <- readRDS(here("output","dada2","seqtable.rds")) # raw sequence table
seqtab.nochim <- readRDS(here("output","dada2","seqtable_nochimeras.rds")) # sequence table w/o chimeras
taxa <- readRDS(here("output","dada2","taxonomic_assignments.rds")) # taxonomic assignments
```

```{r construct phyloseq object from dada2 output}
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(df), 
               tax_table(taxa))
# i skipped the line where Ben removes the mock sample
```

```{r ASV short names}
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps
```


Which ASVs are similarly abundant across sample locations?
Which ASVs differ in abundance across sample locations?

Analysis 1: calculate relative abundance of each ASV within each sample.
Analysis 2: classify types of ASVs to plot with thresholds
Analysis 3: determine change in relative abundance of each ASV between luminal and mucus
Analysis 4: determine change in relative abundance of each ASV between mucus and tissue


```{r analysis1-1}
m1r <- ps.m1 %>%
  # transform abundance from read counts to rel abundance per sample
  tax_glom("Phylum") %>%
  transform_sample_counts(function(OTU) OTU/sum(OTU)) %>%
  psmelt %>%
  ggplot(aes(x=Location, y=Abundance, fill=Phylum)) + 
  geom_col() #+ facet_wrap(~Mouse)
print(m1r)
```

Questions inspired from the above plot:
Are the Cyanobacteria and Verricomicrobiota real? Or contamination?
Are Bacteroidota generally more abundant in crypts/tissue than Firmicutes? Or is this noise from low read count?


```{r melt ps object into dataframe}
dfasv <- ps %>%
#  tax_glom("Family") %>%
  psmelt %>%
  with_groups(Sample, mutate, Proportion = Abundance / sum(Abundance))
```

```{r analysis1.2}
dfasv %>%
  filter(Replicate != 1) %>%
  ggplot(aes(x=Mouse, y=Abundance, fill= Genus)) + 
  geom_col() + facet_wrap(~Location,scales = 'free') +
  theme(legend.position = 'none')
```

```{r analysis1.3}
dfasv %>%
  filter(Replicate != 1) %>%
  ggplot(aes(x=Mouse, y=Proportion, fill= Genus)) + 
  geom_col() + facet_wrap(~Location, scales = 'free_x') +
  theme(legend.position = 'none')
```
Fairly high proportion of each sample are ASVs of unspecified genera.
First thought: continue analysis at ASV level and create a threshold without filtering for ASVs with assignments.
Secondary thought: how many ASVs get assigned to the same genera? Same species (this would require running a different assignment than that in the DADA2 tutorial)?

We don't want to plot 349 ASVs
Instead, filter ASVs that get plotted by the following rules:
   1. ASV must represent 0.1% of community (proportion in samples when present > 0.001)
   2. ASV must change by one standard deviation = 20% of ASV mean proportion (across samples)
   3. ASV must change in same direction (+/-) across all sample pairs

 - Present but noisy ASVs are those that satisfy #1 & 2 but not #3.
 - Present but no change are those that satisfy #1 but not #2 or 3.
 - Lowly abundant ASVs don't even satisfy #1.
 
# First try! ASV grouping across all samples

```{r analysis2 classify ASVs with thresholds}
thresh1 <- dfasv %>%
  with_groups(OTU, mutate, mean_Proportion = mean(Proportion)) %>% 
  filter(mean_Proportion > 0.001) 
```

```{r}
thresh1 %>%
  ggplot(aes(x=Mouse, y=Proportion, color= Genus)) + 
  geom_point() + facet_wrap(~Location,scales = 'free_x') +
  ylim(0, 0.005) +
  theme(legend.position = 'none')
```
```{r}
thresh2 <- thresh1 %>%
  filter(Replicate != 1) %>%
#  tax_glom("Family") %>%
  with_groups(OTU, mutate, SD_asv = sd(Proportion))

thresh2 <- thresh2 %>% 
  filter(SD_asv > 0.05)
```


```{r analysis2 luminal vs mucus, all 3 pairs}
thresh2 %>%
  filter(Location %in% c('Luminal', 'Mucus'), Replicate != 1) %>%
  ggplot(aes(Location, Proportion, color = Family)) +
  geom_point() +
  geom_line(aes(group = Mouse)) +
  facet_wrap(~OTU, scales = 'free_y') +
  theme(legend.position = 'none')
```

```{r analysis3.0 mucus vs tissue, 6 pairs}
thresh2 %>%
  filter(Location %in% c('Mucus', 'Tissue'), Replicate != 1) %>%
  ggplot(aes(Location, Proportion, color = Family)) +
  geom_point() +
  geom_line(aes(group = Mouse)) +
  facet_wrap(~OTU, scales = 'free_y') +
  theme(legend.position = 'none')
```
```{r analysis3.1 mucus vs. tissue, 3 gentle pairs}
thresh2 %>%
  filter(Location %in% c('Mucus', 'Tissue'), Extraction == "2021-07-23", Replicate != 1) %>%
  ggplot(aes(Location, Proportion, color = Family)) +
  geom_point() +
  geom_line(aes(group = Mouse)) +
  facet_wrap(~OTU, scales = 'free_y') +
  theme(legend.position = 'none')
```




# Try 2! Grouping by pairs

Try 2 strategy:
with_group: mouse -> calculate change per ASV in (1) luminal-mucus and (2) mucus-tissue, store new columns
then plot by columns based on filtering by value of calculated change

```{r analysis2 threshold 1, try2}

```


# Oldies from phyloseq_fam
```{r analysis3.2 mucus vs tissue, 3 hard pairs}
thresh2 %>%
  filter(Location %in% c('Mucus', 'Tissue'), Extraction == "2021-07-30") %>%
  ggplot(aes(Location, Proportion, color = Family)) +
  geom_point() +
  geom_line(aes(group = Mouse)) +
  facet_wrap(~OTU, scales = 'free_y') +
  theme(legend.position = 'none')
```
Do trends between gentle or harder scraping preps change? What could lead to change?
1. Decrease suggests more mucus (or tissue) was removed and the bacteria along with it
2. Increase suggests reduction in mucus (or tissue) made remaining bugs more abundant
3. More noise suggests additional changes in direction that were not consistent with previous trend.

Example: [3,3] Erysipelotrichaeceae was increased in tissue from gentle scrapings but mixed (increase, decrease and no change) in hard scraping samples.
It is a quite abundant family, making up ~10% of ASVs in one particular tissue sample. 5-15% of ASVs in the three luminal samples.

- could looking beyond family clear up some of the noise?


Increased:
  1. [1,2] Acholeplasmataceae
  2. [1,4] Akkermansiaceae
  3.
  4. [4,4] Muribaculaceae
  
  changed between analysis 3.1 and 3.2
  5. [5,4] Rikenellaceae
  6. [5,6] Sutterellaceae

Decreased:
  1. 
  
  changed between analysis 3.1 and 3.2
  2. [5,2] Prevotellaceae

No change/not really there:
  1. [1,1] Coprostanoligenes
  2. [1,6] Bacillaceae
  3. [2,2] Bifidobacteriaceae
  4. [2,3] Butyricicoccaceae
  5. [2,4] Christensenellaceae
  6. [3,1] Eggerthellaceae
  7. [3,2] Enterobacteriaceae
  8. [3,4] Fusobacteriaceae
  9. [3,5] Lachnospiraceae
  10. [3,6] Lactobacillaceae
  11. [3,7] Methanobacteriaceae
  12. [4,1] Microbacteriaceae
  13. [4,2] Mitochondria
  14. [4,3] Monoglobaceae
  15. [5,1] Planococcaceae
  16. [5,3] Pseudomonadaceae
  17. 
  18. [5,7] UCG-010
  19. [6,2] Xanthobacteraceae

Present, but noisy:
  1. [1,3] Acidaminococcaceae
  2. [1,7] Bacteroidaceae
  3. [1,8] Barnesiellaceae
  4. [2,5] Clostridiaceae
  5. [2,6] Comanmonadaceae
  6. [2,7] Defluviitaleaceae
  7. [4,5] Oscillospiraceae
  8. [4,6] Peptococcaceae
  9. [4,7] Peptostreptococcaceae
  10. 
  11. 
  12. [5,5] Ruminococcaceae
  13. [6,1] Veillonellaceae
  
  changed between 3.1 and 3.2
  14. [3,3] Erysipelotrichaeceae
  15. [1,5] Anaerovoracaceae



```{r alpha diversity}
plot_richness(ps, x="sample.names", measures=c("Shannon", "Simpson"), color="Location")
```
Harder scrapings tend to have higher diversity in tissue than in mucus.

RE: the message from phyloseq when running this data:
"the data you have provided does not have any singletons. This is highly suspicious. Results of richness
estimates (for example) are probably unreliable, or wrong, if you have already trimmed low-abundance taxa from the data."

Outputs from DADA2 are unlikely to have singletons because single reads are often considered as error.
Especially single reads that are very similar to another more common read, DADA2 often interprets this as sequencing error.
Were we to expect some very lowly abundant organisms in some samples, we could consider a technique called "pseudopooling", which considers whether the single reads are found in different samples on the same plate. This suggests they are true sequence variants!

Overall, considering singletons only matters if we want to interpret them as reads.


```{r Bray Curtis}
# Transform data to proportions as appropriate for Bray-Curtis distances
ps.prop <- transform_sample_counts(ps, function(otu) otu/sum(otu))
ord.nmds.bray <- ordinate(ps.prop, method="NMDS", distance="bray")
```

```{r}
# Bray-Curtis ordination
plot_ordination(ps.prop, ord.nmds.bray, color="Location", title="Bray NMDS")
```

Luminal samples cluster together, and controls (mock, extraction and water) are grouped together away from most mouse samples.
Tissue and mucus samples are not easily distinguished however.
Which points belong to which pairs?
Does one method of scraping group in pairs and one method seem affect their likeness?




# scratch work



```{r}
sessionInfo()
```

